# Matrix Approach to Simple Linear Regression Analysis

## Matrices

- Definition
	- matrix
	- elements
	- dimension
- Notation: a boldface symbol, such as **A**, **X** or **Z**.
- **Square Matrix**: the number of rows equals the number of columns
- **Vector**
	- column vector/vector: only one column matrix
	- row vector
- **Transpose**: **A'** (A prime) is the transpose of a matrix **A**
- **Equality of Matrices**: same dimension and all same corresponding elements
- **design matrix**

## Matrix Addition and Subtraction

- same dimension
- the sum or difference of the corresponding elements of the two matrixs
- **A** + **B** = **B** + **A**

## Matrix Multiplication

- Multiplication of a Matrix by a **scalar**
	- a scalar is an ordinary number or a symbol representing a number
- Multiplication of a Matrix by a Matrix
	- the product **AB**, we say that **A** is postmultiplied by **B** or **B** is premultiplied by **A**
	- $AB \neq BA$

In general, if **A** has dimension r * c and **B** has dimension c * s, the product **AB** is a matrix of dimension r * s, which is

$$AB_{r \times s} = \begin{bmatrix}\sum_{k=1}^{c} a_{ik}b_{kj}\end{bmatrix}\text{, where }i=1,...,r;j=1,...,s$$

- Regression Examples

$$Y'Y_{1 \times 1} = \begin{bmatrix} Y_{1} & Y_{2} & ... & Y_{n} \end{bmatrix}\begin{bmatrix}Y_{1} \\ Y_{2} \\ ... \\ Y_{n} \end{bmatrix} =  Y_{1}^{2} + Y_{2}^{2} + ... + Y_{n}^{2} = \sum Y_{i}^{2}$$ 

$$X'X_{2 \times 2} = \begin{bmatrix} 1 & 1 & ... & 1 \\ X_{1} & X_{2} & ... & X_{n} \end{bmatrix}\begin{bmatrix} 1 & X_{1} \\ 1 & X_{2} \\ ... \\ 1 & X_{n} \end{bmatrix} =  \begin{bmatrix} n & \sum X_{i} \\ \sum X_{i} & \sum X_{i}^{2} \end{bmatrix}$$ 

$$X'Y_{2 \times 1} = \begin{bmatrix} 1 & 1 & ... & 1 \\ X_{1} & X_{2} & ... & X_{n} \end{bmatrix}\begin{bmatrix} Y_{1} \\ Y_{2} \\ ... \\ Y_{n} \end{bmatrix} =  \begin{bmatrix} \sum Y_{i} \\ \sum X_{i}Y_{i} \end{bmatrix}$$ 

## Special Types of Matrices

- **Symmetric Matrix**: **A** = **A'**
	- Symmetric matrix is necessarily square
	- premultiply a matrix by its transpose, say **X'X** is symmetric
- **Diagonal Matrix**: off-diagonal elements are all zeros
- **Identity Matrix**, denoted by **I**: a diagonal matrix whose elements on the main diagonal are all 1s.
	- **AI** = **IA** = **A**, $A, I \in \mathbb{R}^{r \times r}$
- **Scalar Matrix**: a diagonal matrix whose main-diagonal elements are the same, can be expressed as k**I**
- Vector and matrix with all elements unity
	- a column vector with all elements 1 will be denoted by **1**
	- a square matrix with all elements 1 will be denoted by **J**
	- **1'1** = n
	- **11'** = **J**
- Zero Vector: a vector containing only zeros, denoted by **0**

## Linear Dependence and Rank of Matrix

- Linear dependence

We define the set of c column vectors $C_{1}, ..., C_{c}$ in an r $\times$ c matrix to be **linearly dependent** if one vector can be expressed as a linear combination of others. If no vector in the set can be so expressed, we define the set of vectors to be **linearly independent**.

In a more general, when c scalars $k_{1},...,k_{c}$, not all zero, can be found such that:

$$k_{1}\mathbf{C_{1}} + k_{2}\mathbf{C_{2}} + ... + k_{c}\mathbf{C_{c}} = \mathbf{0}$$

where **0** denotes the zero column vector, **the c column vectors** are **linearly dependent**. If the only set of scalars for which the equality holds is $k_{1} = k_{2} = ... = k_{c} = 0$, **the set of c column vectors** is **linearly independent**.

- Rank of Matrix: the maximum number of linearly independent columns in the matrix
	- the rank of r $\times$ c matrix cannot exceed min(r, c)
	- When a matrix is the product of two matrixs, its rank cannot exceed the smaller of the two ranks for the matrices being multiplied.

	
## Inverse of a Matrix

The inverse of a matrix $\mathbf{A}$ is another matrix, denoted by $\mathbf{A^{-1}}$, such that:

$$\mathbf{A}^{-1}\mathbf{A} = \mathbf{AA}^{-1} = \mathbf{I}$$

where **I** is the identity matrix.

- the inverse of a matrix is defined only for square matrix
- many square matrix do not have inverse
- the inverse of a square matrix, if exits, is unique

Finding the inverse

- An inverse of a square $r \times r$ matrix exists if the rank of the matrix is r. Such a matrix is said to be **nonsingular** or of **full rank**.
- An $r \times r$ matrix with rank less than r is said to be **singular** or **not of full rank**, and does not have an inverse.
- The inverse of an $r \times r$ matrix of full rank also has rank r.

If:

$$\mathbf{A}_{2 \times 2} = \begin{bmatrix} a & b \\ c & d \end{bmatrix}$$

then:

$$\mathbf{A}_{2 \times 2}^{-1} = \begin{bmatrix} \frac{d}{D} & \frac{-b}{D} \\ \frac{-c}{D} & \frac{a}{D} \end{bmatrix}$$

where:

$$D = ad - bc$$

D is called the **determinant**(行列式) of the matrix **A**. If **A** were singular, its determinant would equal zero and no inverse of A would exist.

Regression Example

$$\mathbf{X}'\mathbf{X}_{2 \times 2} = \begin{bmatrix} n & \sum X_{i} \\ \sum X_{i} & \sum X_{i}^{2} \end{bmatrix}$$ 

$$ a = n, b = c = \sum{X_{i}}, d = \sum{X_{i}^{2}} $$

$$ D = n\sum{X_{i}^{2}} - (\sum{X_{i}})^{2} = n\sum{(X_{i} - \bar{X})}^{2}$$

$$(\mathbf{X}'\mathbf{X})_{2 \times 2}^{-1} = \begin{bmatrix} \frac{1}{n} + \frac{\bar{X}^2}{\sum(X_{i} - \bar{X})^{2}} & \frac{-\bar{X}}{\sum{(X_{i} - \bar{X})^2}} \\ \frac{-\bar{X}}{\sum{(X_{i} - \bar{X})^2}} & \frac{1}{\sum{(X_{i} - \bar{X})^2}} \end{bmatrix}$$

## Some Basic Results for Matrics

- A + B = B + A
- (A + B) + C = A + (B + C)
- (AB)C = A(BC)
- C(A + B) = CA + CB
- k(A + B) = kA + kB
- (A')' = A
- (A + B)' = A' + B'
- (AB)' = B'A'
- (ABC)' = C'B'A'
- $(AB)^{-1} = B^{-1}A^{-1}$
- $(ABC)^{-1} = C^{-1}B^{-1}A^{-1}$
- $(A^{-1})^{-1} = A$
- $(A')^{-1} = (A^{-1})'$

## Random Vectors and Matrices

- **A random vector or matrix** contains elements that are random variables.
- **Expectation** of random vector or matrix: 

$$\mathbf{Y}_{3 \times 1} = \begin{bmatrix} Y_1 \\ Y_2 \\ Y_3 \end{bmatrix}\text{, and } \mathbf{E(Y)}_{3 \times 1 } = \begin{bmatrix} E(Y_1) \\ E(Y_2) \\ E(Y_3) \end{bmatrix}$$

- **Variance-Covariance Matrix of Random Vector**

$$\sigma^2(\mathbf{Y})_{n \times n} = \begin{bmatrix} \sigma^2(Y_1) & \sigma(Y_1,Y_2) & ... & \sigma(Y_1, Y_n) \\  \sigma(Y_2,Y_1) & \sigma^2(Y_2) & ... & \sigma(Y_2, Y_n) \\ ... & ... & ... & ...\\  \sigma(Y_n, Y_1) &  \sigma(Y_n, Y_2) & ... & \sigma^2(Y_n) \end{bmatrix}$$

which is a symmetric matrix.

- Some Basic Rules

$$\mathbf{W} = \mathbf{AY}$$, 

which **W** and **Y** are two random vectors and **A** is a constant matrix

$$E(\mathbf{A}) = \mathbf{A}$$

$$E(\mathbf{W}) = E(\mathbf{AY}) = \mathbf{A}E(\mathbf{Y})$$

$$\sigma^2(\mathbf{W}) = \sigma^2(\mathbf{AY}) = \mathbf{A}\sigma^2(\mathbf{Y})\mathbf{A'}$$

- Multivariate Normal Distribution


## Simple Linear Regression Model in Matrix Terms

The normal error regression model in matrix terms is:

$$\underset{n \times 1}{\mathbf{Y}} = \underset{n \times 2}{\mathbf{X}}\underset{2 \times 1}{\boldsymbol{\beta}} + \underset{n \times 1}{\boldsymbol{\varepsilon}}$$
, where

$$\underset{n \times 1}{\mathbf{Y}} = \begin{bmatrix}y_1 \\ y_2 \\ ... \\y_n\end{bmatrix}, \underset{n \times 2}{\mathbf{X}} = \begin{bmatrix}1 & x_1 \\ 1 & x_2 \\ ...  & ... \\ 1  & x_n \end{bmatrix}, \underset{2 \times 1}{\boldsymbol{\beta}} = \begin{bmatrix}\beta_0 \\ \beta_1 \end{bmatrix} and \underset{n \times 1}{\boldsymbol{\varepsilon}} = \begin{bmatrix}\varepsilon_1 \\ \varepsilon_2 \\ ... \\\varepsilon_n\end{bmatrix}$$

## Leasst Squares Estimation of Regression Parameters

$$\mathbf{b} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y},$$

where **b** is the vector of the least squares regression coefficients: 

$$ \mathbf{b} = \begin{bmatrix} b_0 \\ b_1 \end{bmatrix} $$

Note that, the inverse is only valid for square matrix and $\mathbf{X}'\mathbf{X}$ is definitely a square matrix.

## Fitted Values and Residuals

$$\hat{\mathbf{Y}} = \mathbf{X}\mathbf{b} = \mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}$$

or

$$\hat{\mathbf{Y}} = \mathbf{H}\mathbf{Y}\text{, with }\underset{n \times n}{\mathbf{H}} = \mathbf{X}(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'$$

The matrix **H** is called **hat matrix**. And it's symmetric and has the special property (called idempotency):

$$\mathbf{HH} = \mathbf{H}$$

$$\mathbf{e} = \mathbf{Y} - \mathbf{\hat{Y}} = \mathbf{Y} - \mathbf{HY} = (\mathbf{I}-\mathbf{H})\mathbf{Y}$$

and the matrix $\mathbf{I}-\mathbf{H}$, like the matrix $\mathbf{H}$, is symmetric and idempotent.

The variance-covariance matrix of residuals **e**:

$$\sigma^2(\mathbf{e}) = \sigma^2 \times (\mathbf{I}-\mathbf{H})$$

and is estimated by:

$$s^2(\mathbf{3}) = MSE \times (\mathbf{I}-\mathbf{H})$$

- Proof:

$$\sigma^2(\mathbf{e}) = \sigma^2((\mathbf{I}-\mathbf{H})\mathbf{Y}) = (\mathbf{I}-\mathbf{H})\times \sigma^2(\mathbf{Y}) \times (\mathbf{I}-\mathbf{H})'$$

$$\sigma^2(\mathbf{Y})= \sigma^2 \times \mathbf{I}$$

$$(\mathbf{I}-\mathbf{H})' = (\mathbf{I}-\mathbf{H})$$

$$(\mathbf{I}-\mathbf{H})(\mathbf{I}-\mathbf{H}) = \mathbf{I}-\mathbf{H}$$


$$\sigma^2(\mathbf{e}) = \sigma^2 \times (\mathbf{I}-\mathbf{H})$$

## Analysis of Variance Results

$$SSTO = \sum(Y_i - \bar{Y})^2 = \sum Y_i^2 - \frac{(\sum Y_i)^2}{n} = \mathbf{Y}'\mathbf{Y} - (\frac{1}{n})\mathbf{Y}'\mathbf{JY}$$

$$SSE = \mathbf{e}'\mathbf{e} = (\mathbf{Y}-\mathbf{Xb})'(\mathbf{Y}-\mathbf{Xb})=\mathbf{Y}'\mathbf{Y} - \mathbf{b}'\mathbf{X}'\mathbf{Y}$$

$$SSR = \mathbf{b}'\mathbf{X}'\mathbf{Y} - (\frac{1}{n})\mathbf{Y}'\mathbf{JY}$$
A quadratic form is defined as:

$$\underset{1 \times 1}{\mathbf{Y}'\mathbf{AY}} = \displaystyle\sum_{i=1}^{n}\displaystyle\sum_{j=1}^{n}a_{ij}Y_iY_j\text{, where }a_{ij} = a_{ji}$$

**A** is a symmetrc n by n matrix and is called **the matrix of the quadratic form**.

So the sums of squares as quadratic forms as follows:

$$SSTO = \mathbf{Y}'[\mathbf{I} - \frac{1}{n}\mathbf{J}]\mathbf{Y}$$

$$SSE = \mathbf{Y}'[\mathbf{I} - \mathbf{H}]\mathbf{Y}$$

$$SSR = \mathbf{Y}'[\mathbf{H} - \frac{1}{n}\mathbf{J}]\mathbf{Y} $$

Quadratic forms play an important role in statistics because all sums of squares in the analysis of variance for linear statistical models can be expressed as quadratic forms.

## Inferences in Regeression Analysis

- Regression Coefficients

The variance-covariance matrix of **b**:

$$\sigma^2(\mathbf{b}) = \begin{bmatrix} \sigma^2(b_0) & \sigma(b_0,b_1) \\ \sigma(b_0,b_1) & \sigma^2(b_1) \end{bmatrix} = \sigma^2 \times (\mathbf{X}'\mathbf{X})^{-1} = \begin{bmatrix} \frac{\sigma^2}{n} + \frac{\sigma^2\bar{X}^2}{\sum(X_i - \bar{X})^2} & \frac{-\bar{X}\sigma^2}{\sum(X_i - \bar{X})^2} \\ \frac{-\bar{X}\sigma^2}{\sum(X_i - \bar{X})^2} & \frac{\sigma^2}{\sum(X_i - \bar{X})^2} \end{bmatrix}$$

And the estimated variance-covariance matrix of **b**, denoted by $s^2(\mathbf{b})$:

$$s^2(\mathbf{b}) = MSE \times  (\mathbf{X}'\mathbf{X})^{-1}$$

- Mean Response

$$s^2(\hat{Y}_h) = MSE \times (\mathbf{X}_{h}'(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}_h) = MSE \times [\frac{1}{n} + \frac{(X_h - \bar{X})^2}{\sum(X_i - \bar{X})^2}]$$

- Prediction of new observation

$$s^2(pred) = MSE \times (1+\mathbf{X}_{h}'(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}_h)$$

## R code

- example data

```{r}
head(trees)
fit_lm = lm(trees$"Girth" ~ trees$"Height")
summary(fit_lm)
fitted(fit_lm)
anova(fit_lm)
```

- Least squaress estimation

```{r}
y = trees$"Girth"
x = cbind(1,trees$"Height")
b = solve(t(x) %*% x) %*% t(x) %*% y
b
```

- fitted values

```{r, cache=TRUE}
h = x %*% solve(t(x) %*% x) %*% t(x)
dim(h); h[1:4,1:4]
yhat = h %*% y
head(yhat)
```

- sum of squares

```{r, cache=TRUE}
SSTO = t(y) %*% y - 1 / length(y) * t(y) %*% matrix(1, nrow=length(y), ncol=length(y)) %*% y
SSTO
```